Efficiently stable presentations from error-correcting codes

### 2023-08-03 efficient_stability.tex

Given an $r$-local $M$ for $\code$, there is a natural \hnote{should be $r$-local, right?} $rq$-local tester $M'$ for $\code'$ which can be described as follows. Index coordinates of $\code'$ by pairs $(i,\alpha)\in [n]\times\F_2^t$, fixing a bijection between $[qn]$ and $[n]\times \F_2^t$. Then $\nu'$ is the uniform mixtures of two distributions, $\nu'_1$ and $\nu'_2$, defined as follows.

To sample from $\nu'_1$, sample $S\sim\nu$ and $\gamma\in \F_q$ uniformly at random. If $M$ performed the check $v_S \cdot u_S=0$, then perform the check 
\begin{align*}
0 &= \tr\big(( v_S \cdot u_S) \gamma\big)\\
&= \sum_{j\in S} \tr( v_{j} u_j \gamma) \\
&= \sum_{j\in S} \kappa(\gamma v_{j}) \cdot \kappa(u_j)\;.
\end{align*}
This check can be performed by evaluating $u$ at the coordinates $(j,\kappa(\gamma  v_{j}))$ for $j\in S$ and summing the results in $\F_2$.

To sample from $\nu'_2$, sample $i\sim[n]$ uniformly at random and $x,y\in\F_2^t$ uniformly at random, and return $\{i\}\times\{x,y,x+y\}$. Execute the tester $M_\had$ for the Hadamard code, i.e.\ check if the three corresponding entries of $u$ sum to zero.

\hnote{added the parity check version of the tester: } Let $M = (h,\nu)$ denote an $r$-local tester for $\code$, where $h \in \F_q^{\ell \times n}$ is a parity check matrix for $\code$. We now describe an $\max\{r,3\}$-local tester $M' = (h',\nu')$ for $\code'$. Define $h'$ to be the block matrix $h'=\begin{pmatrix} A \\ B \end{pmatrix}$ where 
\begin{itemize}
	\item $A \in \F^{nq^2 \times nq}$ is itself a block-diagonal matrix where the diagonal blocks are the $q^2 \times q$ parity check matrix for the Hadamard code. 
	In other words, $A$ can be viewed as $I_{n\times n} \otimes h_\had$ where $h_\had \in \F_2^{q^2 \times q}$ is the parity check matrix for the Hadamard code. 
	\item $B \in \F^{\ell q \times nq}$ is viewed as having rows indexed by pairs $(p,\gamma) \in \{1,\ldots,\ell\} \times \F_q$ and columns indexed by pairs $(i,x) \in \{1,\ldots,n\} \times \F_2^t$. The entry in row $(p,\gamma)$ and column $(i,x)$ is $1$ if and only if $h_{pi} \neq 0$ and $x = \kappa(\gamma h_{pi})$.
\end{itemize}
It is easy to check that $h'$ is a parity check matrix for $\code'$. Furthermore, every row of $h'$ has at most $\max \{r,3\}$ nonzero entries; this is because the rows of $A$ have at most $3$ nonzero entries and the rows of $B$ have at most $r$ nonzero entries. Finally, define the distribution $\nu'$ to be choosing a uniformly random row of the $A$ block matrix with probability $1/2$, and choosing a uniformly random row of the $B$ block matrix with probability $1/2$.

\hnote{maybe good to add a sentence or two of intuition about why this tester makes sense...}

\hnote{Also, probably could use another lemma or proposition here arguing that this is in fact a valid tester.}

>>>>

Given an $r$-local $M=(h,\nu)$ for $\code$, there is a natural $r$-local tester $M'$ for $\code'$ which can be described as follows. Index coordinates of $\code'$ by pairs $(i,\alpha)\in [n]\times\F_2^t$, fixing a bijection between $[qn]$ and $[n]\times \F_2^t$.  We describe an $r$-local tester $M' = (h',\nu')$ for $\code'$. Informally, $h'$ contains two type of checks. First, the $A$-checks consist in the repetition of $n$ copies of the checks for the Hadamard code, one for each Hadamard-code encoding of an $\F_q$-symbol from $\code$. Second, the $B$-checks implement the checks of $\code$ specified by $M$, directly on the Hadamard encoding. More precisely, define $h'$ to be the block matrix $h'=\begin{pmatrix} A \\ B \end{pmatrix}$ where 
\begin{itemize}
	\item $A \in \F^{nq^2 \times nq}$ is itself a block-diagonal matrix where the diagonal blocks are the $q^2 \times q$ parity check matrix for the Hadamard code. 
	In other words, $A$ can be viewed as $I_{n\times n} \otimes h_\had$ where $h_\had \in \F_2^{q^2 \times q}$ is the parity check matrix for the Hadamard code. 
	\item $B \in \F^{\ell q \times nq}$ is viewed as having rows indexed by pairs $(p,\gamma) \in \{1,\ldots,\ell\} \times \F_q$ and columns indexed by pairs $(i,x) \in \{1,\ldots,n\} \times \F_2^t$. The entry in row $(p,\gamma)$ and column $(i,x)$ is $1$ if and only if $h_{pi} \neq 0$ and $x = \kappa(\gamma h_{pi})$.
\end{itemize}
Define the distribution $\nu'$ as the uniform mixture of the uniform distribution on the rows of the $A$ block matrix and the uniform distribution on the rows of the $B$ block matrix.

\begin{claim}
$h'$ is a parity check matrix for $\code'$.
\end{claim}

\begin{proof}
Let $x\in \F_2^{qn}$ be such that $h'x=0$. Since the $A$-checks enforce that each block of $k$ symbols contains the Hadamard-code encoding of an $\F_q$ symbol, $x$ can be decoded to $x'\in \F_q^n$ such that for each $(i,x)\in \{1,\ldots,n\}\times \F_2^t$, $x_{(i,x)} = \kappa(x'_i)\cdot x$. If the $p$-th row of $h$ enforces the check $v_p\cdot x'=0$, where $v_p\in \F_q^n$, then 
the $(p,\gamma)$-th row of $B$ enforces the check 
\begin{align*}
0&=\sum_{j=1}^n \kappa(\gamma (v_p)_j) \cdot \kappa(x'_j)\\
&= \sum_{j=1}^n \tr( \gamma (v_p)_j x'_j) \\
&= \tr(\gamma (v_p\cdot  x')\;.
\end{align*}
Therefore, $h'x=0$ is equivalent to $\tr(\gamma(v_p \cdot x'))=0$ for all $\gamma$, which is equivalent to $v_p\cdot x'=0$. Thus $\code'=\ker h'$, as desired. 
\end{proof}

\hnote{maybe good to add a sentence or two of intuition about why this tester makes sense...}\tnote{added}

\hnote{Also, probably could use another lemma or proposition here arguing that this is in fact a valid tester.}\tnote{added above, if this what you meant?}
